---
title: "Coding refresher: R and GitHub"
subtitle: "Data Science for Environmental Policy"
author: "DSEP"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    theme: flatly
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

## Today's destination

> **A note before we begin:** Most coding exercises in this course will be
> messier and much more open ended than this one. This activity is a quick refresher for
> students with coding experience and an introduction for students who have
> not coded much (or at all). Please complete it without an LLM or AI coding assistant.
> We will use those tools throughout the course; today is about being sure you
> can read, write, and troubleshoot the basic building blocks!

By the end of this activity, you will be able to:

- explain the difference between R, an R package, and an R project;
- create and inspect objects, vectors, and data frames;
- filter, summarize, and plot environmental data;
- write and use a small function; and
- save, commit, and push your work to GitHub.

There is no expectation that you remember every command. The goal is to get
comfortable reading code, changing it, running it, and checking the results.

## 1. Open the project

Work in the copy of this repository provided by the course. Open the whole
`coding-intro` project in RStudio, either locally or through a GitHub Codespace,
rather than opening only this file. This makes the repository root your working
directory and gives both setups the same RStudio workflow.

Check where R is currently working:

```{r working-directory}
getwd()
list.files()
```

Avoid putting personal absolute paths such as
`/Users/your-name/Desktop/...` in shared code. Project-relative paths make an
analysis easier for someone else to reproduce.

## 2. R and packages

**R** is the programming language. **Packages** are collections of functions,
data, and documentation that extend R. For example, `ggplot2` provides tools
for plotting and `dplyr` provides tools for transforming tabular data.

Install a package once on a computer:

```{r install-example, eval=FALSE}
install.packages("ggplot2")
```

Load it each time you start a new R session and need it:

```{r packages}
library(dplyr)
library(ggplot2)
```

Use `?function_name` or `help("function_name")` when you need documentation.

```{r help-example, eval=FALSE}
?mean
help("filter", package = "dplyr")
```

> **Check:** Why should `install.packages()` usually not appear in an analysis
> script that you run repeatedly?

## 3. Objects and vectors

R stores values in **objects**. The assignment operator `<-` reads as "gets."

```{r objects}
policy_target <- 15
observed_reduction <- 12.4
target_met <- observed_reduction >= policy_target

policy_target
target_met
```

A **vector** stores values of the same basic type. Many R operations are
vectorized: they apply to every value without an explicit loop.

```{r vectors}
annual_emissions <- c(42.1, 40.8, 39.5, 37.9, 36.4)
years <- 2021:2025

annual_emissions
mean(annual_emissions)
annual_emissions - min(annual_emissions)
```

Try these in the Console first, then put your final code in the chunk below.

```{r vectors-practice, eval=FALSE}
# 1. Create a vector named pm25 with the values 18, 15, 13, 11, and 9.
# 2. Calculate its median.
# 3. Test which values are above the WHO annual guideline of 5 micrograms/m3.
```

## 4. Data frames

A **data frame** is a rectangular table. Columns are variables and rows are
observations. We will use R's built-in `airquality` data, which contains daily
air-quality measurements from New York in 1973.

```{r inspect-data}
air <- airquality

head(air)
glimpse(air)
summary(air)
```

Missing observations are represented by `NA`. Ask explicitly whether a value
is missing with `is.na()`.

```{r missing-data}
sum(is.na(air$Ozone))
mean(air$Ozone, na.rm = TRUE)
```

> **Discuss:** Silently dropping missing observations can change a policy
> conclusion. What would you want to know about why ozone measurements are
> missing before interpreting the mean?

## 5. Transform data with dplyr

The pipe `|>` passes the result on its left into the next function. Read the
following code from top to bottom: start with `air`, keep complete ozone
measurements, add readable month labels, group by month, and calculate summary
statistics.

```{r summarize-data}
monthly_air <- air |>
  filter(!is.na(Ozone)) |>
  mutate(month = factor(
    Month,
    levels = 5:9,
    labels = c("May", "June", "July", "August", "September")
  )) |>
  group_by(month) |>
  summarize(
    days_observed = n(),
    mean_ozone = mean(Ozone),
    max_ozone = max(Ozone),
    .groups = "drop"
  )

monthly_air
```

Now modify the pipeline yourself.

```{r transform-practice, eval=FALSE}
# 1. Start with air.
# 2. Keep rows where Temp is at least 80 degrees Fahrenheit.
# 3. Group by Month.
# 4. Calculate the number of days and mean Wind for each month.
# 5. Store the result as warm_days.
```

## 6. Make a plot

`ggplot2` builds a figure in layers. The data and aesthetic mapping define the
foundation; geometries such as points or columns determine how values appear.

```{r plot, fig.alt="Column chart of mean ozone by month from May to September. July and August have the highest means."}
ggplot(monthly_air, aes(x = month, y = mean_ozone)) +
  geom_col(fill = "#2f6f4e") +
  labs(
    title = "Average observed ozone was highest in midsummer",
    subtitle = "New York air-quality observations, 1973",
    x = NULL,
    y = "Mean ozone (ppb)"
  ) +
  theme_minimal(base_size = 12)
```

> **Interpret:** Write one sentence describing what the figure shows and one
> sentence naming a limitation. Remember that these are observations from one
> place and year, not evidence that the month caused the ozone level.

## 7. Write a function

A function packages a repeatable operation. This function converts Fahrenheit
to Celsius:

```{r functions}
fahrenheit_to_celsius <- function(temperature_f) {
  (temperature_f - 32) * 5 / 9
}

fahrenheit_to_celsius(80)
fahrenheit_to_celsius(air$Temp[1:5])
```

```{r function-practice, eval=FALSE}
# Write percent_change(old_value, new_value), then use it to calculate the
# percentage change in annual_emissions from 2021 to 2025.
```

## 8. Read errors as information

Errors are part of programming. Read from the final line upward and check:

1. Did I spell every object and function name correctly?
2. Did I close every quote and parenthesis?
3. Did I load the package that contains the function?
4. Is the object the type or shape the function expects?

Run this code, read the error, and repair it:

```{r error-practice, eval=FALSE}
mean(annual_emission)
```

## 9. Introduction to Git and GitHub

**Git** records the history of a project on your computer. **GitHub** stores a
copy of that project online so you can back it up, share it, and collaborate.
A project tracked by Git is called a **repository** (or **repo**).

Four actions form the basic workflow:

- **Pull** downloads commits from GitHub to your copy of the repository.
- **Stage** selects the changed files that will go into your next snapshot.
- **Commit** records the staged changes with a short description.
- **Push** uploads your local commits to GitHub.

Saving a file does not commit it, and committing does not push it. A commit
stays on your computer or in your Codespace until you push it.

### Make and inspect a change

Add your name below, then save this file:

**Completed by:** _add your name here_

### Commit and push from the terminal

Open RStudio's terminal with **Tools > Terminal > New Terminal**. Run one
command at a time and inspect its output before continuing:

```{bash git-workflow, eval=FALSE}
# Download any commits that are already on GitHub.
git pull

# List the files you have changed.
git status

# Inspect the exact lines you changed.
git diff

# Stage this workbook for the next commit.
git add coding-intro.qmd

# Confirm what is staged.
git status
git diff --staged

# Record the staged change in the repository history.
git commit -m "Complete R coding refresher"

# Upload the commit to GitHub.
git push
```

After `git push` finishes, open the repository on GitHub and confirm that your
name and commit appear. If Git asks you to sign in, follow the browser prompt
and authorize GitHub Codespaces or RStudio.

Do not commit passwords, access tokens, private data, or large generated files.
Before pushing, use `git status` and `git diff --staged` to verify exactly what
will be shared.

### Your first push

Complete this checklist before moving on:

- [ ] I added my name and completed at least one practice chunk.
- [ ] I used `git diff` to review my changes.
- [ ] I staged `coding-intro.qmd` with `git add`.
- [ ] I committed it with `git commit` and a descriptive message.
- [ ] I pushed the commit with `git push`.
- [ ] I found the commit and my changes on GitHub.

For more practice, complete GitHub Skills'
[Introduction to GitHub](https://github.com/skills/introduction-to-github). It
is a free, guided exercise covering repositories, branches, commits, and pull
requests.

## 10. Exit check

Answer these in your own words:

1. What is the difference between installing and loading an R package?
2. What does the pipe `|>` do?
3. What does `NA` mean, and why does it matter?
4. What is the difference between Git and GitHub?
5. What question or concept should we revisit next class?

## Solutions

Open this section after attempting the exercises.

<details>
<summary>Show suggested code</summary>

```{r solutions, eval=FALSE}
pm25 <- c(18, 15, 13, 11, 9)
median(pm25)
pm25 > 5

warm_days <- air |>
  filter(Temp >= 80) |>
  group_by(Month) |>
  summarize(
    days = n(),
    mean_wind = mean(Wind),
    .groups = "drop"
  )

percent_change <- function(old_value, new_value) {
  (new_value - old_value) / old_value * 100
}

percent_change(annual_emissions[1], annual_emissions[5])

# The object is named annual_emissions, with a final "s".
mean(annual_emissions)
```

</details>

## 11. What if we used Python?

This activity uses R, but we could have completed the same analysis in Python
without major conceptual differences. Both languages let us store values,
transform data, write functions, make figures, and work reproducibly. The
syntax and package names change, but the analytical reasoning does not.

For example, earlier we created a vector of annual emissions and calculated
its mean and its distance from the minimum value:

```{r python-comparison-r, eval=FALSE}
annual_emissions <- c(42.1, 40.8, 39.5, 37.9, 36.4)

mean(annual_emissions)
annual_emissions - min(annual_emissions)
```

Here is the same work in Python using only built-in language features:

```python
annual_emissions = [42.1, 40.8, 39.5, 37.9, 36.4]

mean_emissions = sum(annual_emissions) / len(annual_emissions)
distance_from_minimum = [
  emission - min(annual_emissions)
  for emission in annual_emissions
]

print(mean_emissions)
print(distance_from_minimum)
```

R uses `<-` for assignment and applies subtraction directly across a vector.
Python uses `=` for assignment; here, a list comprehension applies the same
subtraction to every value. In a typical Python data analysis, packages such as
`pandas`, `numpy`, and `matplotlib` provide tools similar to `dplyr`, R vectors,
and `ggplot2`.

Git and GitHub do not depend on the programming language. The same `git pull`,
`git add`, `git commit`, and `git push` commands work for a repository containing
R, Python, or both.